FlowFPX: Nimble Tools for Debugging Floating-Point Exceptions

### 2023-08-17 header.tex
checkpoint: 1,2

\keywords{Julia, Floating point, Numerical computing}

\hypersetup{
pdftitle = {FlowFPX: Nimble Tools for Debugging Floating-Point Exceptions},
pdfsubject = {JuliaCon 2022 Proceedings},
pdfauthor = {Taylor Allred, Xinyi Li, Ashton Wiersdorf, Ben Greenman, Ganesh Gopalakrishnan},
pdfkeywords = {Julia, Floating point, Numerical computing},
}

>>>>

\keywords{Julia, floating-point, debugging}

\hypersetup{
pdftitle = {FlowFPX: Nimble Tools for Debugging Floating-Point Exceptions},
pdfsubject = {JuliaCon 2022 Proceedings},
pdfauthor = {Taylor Allred, Xinyi Li, Ashton Wiersdorf, Ben Greenman, Ganesh Gopalakrishnan},
pdfkeywords = {Julia, floating-point, debugging},
}

### 2023-08-17 paper.tex
fix includegraphics, move figs up
[ht] is not the same as [th]

top figures are best to avoid confusion about boundaries
between text, figure, and captions
checkpoint: 1,2

% Misc typographical tweaks.
\setlength{\parindent}{1.2em}
% \setmonofont{JuliaMono}[Extension = .ttf, Path = ./, UprightFont = *-Regular, ItalicFont = *-Italic]

@@

\begin{abstract}
  Reliable numerical computations are central to high-performance computing and machine learning.
  However, floating-point numbers exhibit non-intuitive behaviors that make writing correct numerical computation difficult.
  In particular, \emph{floating-point exceptions} occur when no mathematically sensible result is available, and it is difficult to find \emph{where} these exceptions occur.
  Unhandled exceptions can cause programs crash, exhibit bad behavior, or worse, silently produce good-looking but incorrect data.
  We present FlowFPX: a tool set for debugging floating-point code.
  FlowFPX can track floating-point exceptions occurring in Julia code to their sources.
  FlowFPX also detects otherwise silent exceptions which can foul program outputs.
  FlowFPX produces intuitive visualizations of summarized exception flows including how they are generated, propagated and killed, thus helping with debugging.
\end{abstract}

@@

Reliable numeric computations are central to high-performance computing,
machine learning, and scientific applications.
Yet, the \fp{} arithmetic that powers these computations is fundamentally
unreliable~(\cref{s:background}).
Exceptional values, such as Not a Number (\Nan{}) and infinity (\Inf{}),
can and often do arise thanks to culprits such as roundoff error,
catastrophic cancellation, singular matrices, and vanishing
derivatives~\cite{sdjmrstp-pc-2022,ddghlllprr-correctness-2022,gllprt-correctness-2021,fpchecker-reports,llg-soap-2022,bllmg-xloop-2022}.
When exceptions happen, domain experts must take time away from their main
focus and trek down a rabbit hole to find the source of the error.
There is little tool support to assist in this difficult task,
thus (unsurprisingly) a quick GitHub search reports over 4,000 open issues
related to \NaN{} or \Inf{} exceptions~\cite{github-issues}.

We introduce \FlowFPX{}, a toolkit for debugging
\fp{} exceptions~(\cref{s:flowfpx}):
\begin{itemize}
  \item
    The centerpiece of \FlowFPX{} is \FT{}, a dynamic analysis tool that uses
    Julia's operator overloading to selectively monitor for exceptions and to
    fuzz code for vulnerabilities.
  \item
    To visualize results, \FT{} adapts coalesced stack-trace graphs 
    (CSTGs, or \CSTG{}s)~\cite{humphreySystematicDebuggingMethods2014} for a whole-program view on the generation,
    propagation, and death of exceptional values.
  \item A companion tool \GPUFPX{}~\cite{llsflg-hpdc-2023} provides fine-grained insights for GPU kernels.
\end{itemize}
%
\FlowFPX{} has helped debug exceptions in a variety
of applications: from ocean simulations to heat models~(\cref{s:casestudies}).
This paper concludes with related work~(\cref{s:related}) and a
brief discussion~(\cref{s:discussion}).

@@

\Fp{} is essentially scientific notation---\fp{} numbers are represented by packing a sign bit, an exponent, and a fraction part (also called the ``significand'' or ``mantissa'') into contiguous bit fields.
The advantages of this representation are that allows for compact representation of both very small and very large numbers in a narrow range of bits---a 64-bit float can represent numbers as big as $2^{1024}$ and as small as $2^{-1074}$.
This range is sufficient to represent numbers as small as Planck's constant or as big as Avogadro's constant with reasonable accuracy.

\begin{figure}[h]
  \label{fig:real_vs_float}
  \includegraphics[width=\columnwidth]{ fig/real_vs_fp.png }
  \caption{The real number line vs the points representable with \fp{}.}
\end{figure}

The flip side is that not every real number within that wide range is representable as a \fp{} value---there is almost always some gap between the real number that we wish to represent and the \fp{} number that approximates that value.
Whenever a value we have doesn't have a precise representation, we must \emph{round} that real value to the nearest \fp{} value, which introduces some error in our calculations.
In addition to this error, the finite nature of \fp{} values means that we can run into some situations that wouldn't arise when working with real numbers.

For example, if two numbers' exponents differ by too large an amount, adding or subtracting the smaller from the bigger will leave us with just the bigger number.
Below we take the Planck constant and add it to Avogadro's constant and are left bit-for-bit with Avogadro's constant, as if we had never done any addition at all: there is no way to pack all the significant figures from true value that differs minutely from Avogadro's number into a 64-bit \fp{} number.

\begin{lstlisting}[language = Julia]
p = 6.62607015e-34
a = 6.02214076e23

println("planck:   $p : $(bitstring(p))")
println("avogadro: $a  : $(bitstring(a))")
println("p + a:    $(p + a)  : $(bitstring(p + a))")
\end{lstlisting}

You can see in the output that \texttt{p + a} is bit-for-bit equivalent to \texttt{a}:

\begin{lstlisting}
planck:   6.62607015e-34 : 0011100100001 ... 001
avogadro: 6.02214076e23  : 0100010011011 ... 111
p + a:    6.02214076e23  : 0100010011011 ... 111
\end{lstlisting}

This is a fairly extreme example; even when the numbers are closer in magnitude, most operations on \fp{} numbers induce loss of accuracy.

\subsection{Floating-Point Standards}

Prior to 1985, incomparable techniques for representing \fp{} values proliferated as mainframes and later personal computers emerged.
This made writing portable \fp{} code challenging.~\cite{houghIEEEStandard7542019}
IEEE 754~\cite{IEEEStandardBinary1985} standardizes \fp{} representation and operations across different architectures---these days, most machines and languages follow the IEEE 754 standard, and we make no effort to generalize our work beyond what IEEE 754 specifies.

\subsection{Exceptions and Exceptional Values}

IEEE 754 defines \emph{\fp{} exceptions}, which are events that occur when something goes wrong in a calculation.
William Kahan says,

\begin{quote}
An arithmetic exception arises when [there is] no result that would be acceptable universally.
\end{quote}

For example, dividing by zero is an exception, as is exceeding a \fp{} format's range.
These are exceptions because it is impossible to say what the right answer \emph{should} have been: perhaps that zero was meant to be a value that got really \emph{close} to zero,\footnote{The third kind of exceptional value is called a \emph{subnormal} value.} and suddenly rounded towards zero before division.
Or maybe that value that got really big \emph{does} need to get that big, but you need to start using a larger bit width for your \fp{} values.
Whatever, the case, there isn't a universally acceptable answer, so a \fp{} exception gets raised.

IEEE 754 doesn't force the programmer to always handle these exceptions, however, so we end up with \emph{exceptional values}, which are the results of exception-raising operations.
The two values we work with in this paper are \Inf{}\footnote{There are actually two infinities: $+\infty$ and $-\infty$; they behave the same for our purposes, so we will refer to them just as ``\Inf{}''.} and \NaN{}, which result from overflow and nonsensical operations respectively.
\Cref{fig:nan-gens} lists some of the situations that lead to exceptional values.

\begin{figure}[ht]
  \centering
  \label{fig:nan-gens}
  % \begin{tabular}{c | p{0.5 \columnwidth}}
  \begin{tabular}{c | c}
    \NaN{} & \Inf{} \\
    \hline
    $\tfrac{0}{0}$ & $\tfrac{x}{0},\, x \neq 0$ \\
    $0 * \infty$ & $\log{(0)}$ \\
    $\text{\texttt{NaN}} + x$ & $\infty + x,\, x \neq \text{\texttt{NaN}}$ \\
    $\infty - \infty$ & Any result exceeding \\
    & representation capacity
  \end{tabular}
  \caption{Examples of operations leading to exceptional values}
\end{figure}

For more details on exceptional values, see~\cite{knuthArtComputerProgramming1997,torontoPracticallyAccurateFloatingPoint2014} as well as the \textit{Handbook of Floating-Point Arithmetic}.~\cite{mullerHandbookFloatingPointArithmetic2018}

\subsection{Lifetimes of Unhandled Exceptions}
\label{s:to-kill-a-fp}

Exceptional values have a life cycle: instead a program-halting exception, an exceptional value comes into being at the point of an undefined operation and propagates through the rest of the computation.
Often they go on to manifest in the output of the code, but sometimes they can ``die'' by disappearing silently.
We characterize this life cycle in \cref{f:gpk}.

\begin{figure}[ht]
  \includegraphics[width=\columnwidth]{fig/genpropkill-outline.png}
  \caption{Gen, Prop, Kill: Lifetime of a \NaN{} value}
  \label{f:gpk}
\end{figure}

When a \NaN{} or an \Inf{} comes into being as the result of an undefined operation, we call this a generation, or \emph{gen} for short.

Most of the time, when any or all of the arguments are \NaN{}, the result is also a \NaN{}.
We call this a propagation, or \emph{prop} for short.
This is good and desirable, because it means that an arithmetic error---even if buried deep in an expression---will still be observable, allowing a user to find and debug it.

However, there are a few places where exceptional values flow into an expression, but not out.
We call these instances \emph{kills}, and these can lead to subtle logic bugs.
For example, \emph{any} comparison ($<$, $>$, $=$, $\neq$, etc.) with a \NaN{} returns \texttt{false}.

@@

\begin{lstlisting}[language = Julia]
function bad_max(lst)
  max_seen = 0.0
  for x in lst
    if ! (x <= max_seen)
      # swap if new val greater
      max_seen = x
    end
  end
  max_seen
end

function good_max(lst)
  foldl(max, lst)
end
\end{lstlisting}

Both \texttt{bad\_max} and \texttt{good\_max} compute the maximum in a list.
However, the \texttt{<=} operator used in the \texttt{bad\_max} implementation does \emph{not} propagate \NaN{}s: \texttt{NaN <= 4} returns \texttt{false}, not \NaN{}.
Because of this, you might not get the result you expected:

\begin{lstlisting}[language = Julia]
bad_max([1, 5, NaN, 4])     # returns 4
good_max([1, 5, NaN, 4])    # returns NaN
\end{lstlisting}

Not only is the result from \texttt{bad\_max} problematic for obscuring the fact that there was a \NaN{} in the list, the result is also wrong!
The built-in \texttt{max} function correctly propagates \NaN{}s, so we can see from its return that we got a \NaN{} somewhere.
It's not entirely obvious \emph{where} that \NaN{} came from.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=3in]{./fig/cstg_static_diagram.png}
  \caption{CSTG makes a graph by connecting adjacent stack frames}
  \label{fig:cstg_demo}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=3in]{./fig/float_tracker_context_boxes.png}
  \caption{FloatTracker and GPU-FPX compliment each other}
  \label{fig:ft_and_gpu_fpx}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=3in]{./fig/shallow_waters_cfl_diff.png}
  \caption{Effect of high CFL parameters: on the left, normal output. On the right: \NaN{}s cause white holes and gaps in the result.}
  \label{fig:sw_nans}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=3in]{fig/sw_nan_cstg_clean.png}
  \caption{CSTG of \NaN{} generation logs from ShallowWaters.}
  \label{fig:sw_nan_cstg}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[width=3in]{fig/sw_inf_cstg_clean.png}
  \caption{CSTG of \Inf{} generation logs from the same run as~\cref{fig:sw_nan_cstg}.}
  \label{fig:sw_inf_cstg}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=3in]{./fig/cstg_diff_pretty.png}
  \caption{Diff between the early and latter stages of ShallowWaters}
  \label{fig:cstg_diff_demo}
\end{figure}

@@

In a similar vein, Herbie\cite{panchekhaAutomaticallyImprovingAccuracy2015} is a tool for automatically rewriting arithmetic expressions to reduce \fp{} error.
\FT{} tracks down exceptional \fp{} values; once the location of a gen is found, Herbie could be employed to rewrite the arithmetic expression, if \fp{} error is the cause of the exceptional value.
Odyssey\cite{misbackOdysseyInteractiveWorkbench2023} provides a helpful interactive interface to the tooling that Herbie provides.

>>>>

% Misc typographical tweaks.
\setlength{\parindent}{10pt}
% \setmonofont{JuliaMono}[Extension = .ttf, Path = ./, UprightFont = *-Regular, ItalicFont = *-Italic]

@@

\begin{abstract}
  Reliable numerical computations are central to scientific computing,
  but the floating-point arithmetic that enables large-scale models
  is fundamentally unreliable.
  Numeric exceptions are a common occurrence and can propagate through
  code, leading to flawed results.
  This paper presents FlowFPX, a toolkit for systematically debugging
  floating-point exceptions by recording their flow,
  coalescing exception contexts, and fuzzing in select locations.
  These tools help scientists discover when exceptions happen
  and track down their origin, smoothing the way to a reliable codebase.
\end{abstract}

@@

Reliable numeric computations are central to high-performance computing,
machine learning, and scientific applications.
Yet the \fp{} arithmetic that powers these computations is fundamentally
unreliable~(\cref{s:background}).
Exceptional values, such as Not a Number (\Nan{}) and infinity (\Inf{}),
can and often do arise thanks to culprits such as roundoff error,
catastrophic cancellation, singular matrices, and vanishing
derivatives~\cite{sdjmrstp-pc-2022,ddghlllprr-correctness-2022,gllprt-correctness-2021,fpchecker-reports,llg-soap-2022,bllmg-xloop-2022}.
Developers are responsible for guarding against exceptions, but this task
is difficult because many operations can generate and propagate exceptions.
Worst of all, operations such as less-than (\code{<}) can kill an exceptional value,
leaving no trace of the problem.
There is little tool support to assist in exception debugging,
thus (unsurprisingly) a quick GitHub search reports over 4,000 open issues
related to \NaN{} or \Inf{} exceptions~\cite{github-issues}.

This paper introduces \FlowFPX{}, a toolkit for debugging
\fp{} exceptions~(\cref{s:flowfpx})
that hase helped improve a variety of applications,
from ocean simulations to heat modes~(\cref{s:casestudies}):
\begin{itemize}
  \item
    The centerpiece of \FlowFPX{} is \FT{}, a dynamic analysis tool that
    selectively monitors for exceptions and fuzzes code for vulnerabilities.
  \item
    To visualize results, \FT{} adapts coalesced stack-trace graphs (CSTGs, or
    \CSTG{}s)~\cite{humphreySystematicDebuggingMethods2014}
    to summarize the program contexts that handled exceptional values.
  \item
    A companion tool \GPUFPX{}~\cite{llsflg-hpdc-2023} provides fine-grained insights for GPU kernels.
\end{itemize}

@@

\begin{figure}[t]\centering
  \includegraphics[trim=0 0 0 60,clip,width=0.95\columnwidth]{fig/real_vs_fp.png}
  \caption{Floats are spread across the real number line.}
  \label{f:real-vs-fp}
\end{figure}

\Fp{} numbers use a finite number of bits to represent a spectrum of points along the real number line~(\cref{f:real-vs-fp}).
The implementation strategy is essentially scientific notation.
A \fp{} number packs a sign bit, an exponent, and a fraction part (also called the ``significand'' or ``mantissa'') into a bitstring (typically 64 or 32 bits, but 16-bit and 8-bit formats are on the rise~\cite{klowerLowprecisionClimateComputing2021,fp8}).
This representation supports very small and very large numbers in a narrow range of bits.

\begin{lstlisting}[language = Julia]
julia> prevfloat(typemax(Float64))
1.7976931348623157e308
julia> nextfloat(typemin(Float64))
-1.7976931348623157e308
\end{lstlisting}

The flip side is that most real numbers fall into the gaps between \fp{} numbers and must be rounded, which introduces error
and can lead to surprising results.
For example, adding the tiny Planck constant to the large Avogadro number results
in Avogadro's number after rounding:

\begin{lstlisting}[language = Julia]
julia> planck = 6.62607015e-34
6.62607015e-34
julia> avogadro = 6.02214076e23
6.02214076e23
julia> avogadro + planck == avogadro
true
\end{lstlisting}

This is an extreme example, but many operations on \fp{} numbers closer in magnitude induce a loss of accuracy.
Refer to the literature for more details, e.g.,~\cite{knuthArtComputerProgramming1997,torontoPracticallyAccurateFloatingPoint2014,mullerHandbookFloatingPointArithmetic2018}.

\subsection{Exceptions and Exceptional Values}
\label{s:exnvalue}

The IEEE~754 \fp{} standard~\cite{IEEEStandardBinary1985}
defines exceptions and exceptional values as the outcome of
operations that have ``no single universally acceptable result''~\cite{p-draft-1997}.
For example, dividing by zero and exceeding the \code{Float64} range
both lead to exceptions:

% julia> 2 / 0
% Inf
\begin{lstlisting}[language = Julia]
julia> 0 / 0
NaN
julia> avogadro^avogadro
Inf
julia> log(0)
-Inf
julia> Inf + NaN
NaN
\end{lstlisting}

IEEE~754 defers the question of how to handle such exceptions to application code.
It is up to developers to watch for \Nan{} and \Inf{} (and also subnormal numbers)
and implement an appropriate repair.
Without guards, bugs happen.

\subsection{Lifetime of an Unhandled Exceptions}
\label{s:to-kill-a-fp}

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{fig/genpropkill-outline.png}
  \caption{Gen, Prop, Kill: Lifetime of a \NaN{} value}
  \label{f:gpk}
\end{figure}

Exceptional values have a lifetime: they are born, or \emph{generated}, by some operation; they \emph{propagate} through other operations; and they either appear in the program output, go out of scope, or get \emph{killed} by a numeric operation.
\Cref{f:gpk} summarizes this \emph{gen}-\emph{prop}-\emph{kill} process.
The gens and props are straightforward; see above for examples~(\cref{s:exnvalue}).
The kills often arise from numeric comparisons (\code{<}, \code{=}, etc.),
but exponents (\code{1\^NaN}) and over-eager matrix optimizations can kill
exceptions~\cite{ddghlllprr-correctness-2022}.

To illustrate TODO

@@

\begin{lstlisting}[language = Julia]
function max1(lst)
  max_seen = 0.0
  for x in lst
    # swap if x is not too small
    if ! (x <= max_seen)
      max_seen = x
    end
  end
  max_seen
end

function max2(lst)
  foldl(max, lst)
end
\end{lstlisting}

Both \code{max1} and \code{max2} compute the maximum in a list.
However, the \texttt{<=} operator used in the \code{max1} implementation does \emph{not} propagate \NaN{}s: \texttt{NaN <= 4} returns \texttt{false}, not \NaN{}.
Because of this, you might not get the result you expected:

\begin{lstlisting}[language = Julia]
max1([1, 5, NaN, 4])     # returns 4
max2([1, 5, NaN, 4])    # returns NaN
\end{lstlisting}

Not only is the result from \texttt{max1} problematic for obscuring the fact that there was a \NaN{} in the list, the result is also wrong!
The built-in \texttt{max} function correctly propagates \NaN{}s, so we can see from its return that we got a \NaN{} somewhere.
It's not entirely obvious \emph{where} that \NaN{} came from.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=3in]{./fig/cstg_static_diagram.png}
  \caption{CSTG makes a graph by connecting adjacent stack frames}
  \label{fig:cstg_demo}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=3in]{./fig/float_tracker_context_boxes.png}
  \caption{FloatTracker and GPU-FPX compliment each other}
  \label{fig:ft_and_gpu_fpx}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=3in]{./fig/shallow_waters_cfl_diff.png}
  \caption{Effect of high CFL parameters: on the left, normal output. On the right: \NaN{}s cause white holes and gaps in the result.}
  \label{fig:sw_nans}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=3in]{fig/sw_nan_cstg_clean.png}
  \caption{CSTG of \NaN{} generation logs from ShallowWaters.}
  \label{fig:sw_nan_cstg}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=3in]{fig/sw_inf_cstg_clean.png}
  \caption{CSTG of \Inf{} generation logs from the same run as~\cref{fig:sw_nan_cstg}.}
  \label{fig:sw_inf_cstg}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=3in]{./fig/cstg_diff_pretty.png}
  \caption{Diff between the early and latter stages of ShallowWaters}
  \label{fig:cstg_diff_demo}
\end{figure}

@@

In a similar vein, Herbie\cite{panchekhaAutomaticallyImprovingAccuracy2015} is a tool for automatically rewriting arithmetic expressions to reduce \fp{} error.
\FT{} tracks down exceptional \fp{} values; once the location of a gen is found, Herbie could be employed to rewrite the arithmetic expression, if \fp{} error is the cause of the exceptional value.
The Odyssey\cite{misbackOdysseyInteractiveWorkbench2023} workbench provides an interactive interface to Herbie.
